VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-16 doc/publications/journal2014/paper.tex
remove VCF format

\subsection*{ADMIXTURE Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from the VCF file, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.

%% TODO - NEIL - (and better heading) %%
\subsection*{Personal Genome Project Data}
%% The PGP data..... 
PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
file using VCFtools \cite{Danecek2011Variant}.

@@

%Hadoop
%1. Map
%The mapper converts each line of the VCF file into a preliminary key-value pair, where the Key is the byte-position of the first character thereby functioning as a proxy for the genomic location (variant ID), and the Value is a string of the line. From this, the key-value pairs for the individual genotypes are created by extracting the variant from each sample as the new Value. The new key is a tuple consisting of the primary key, which is the individual ID (position of the parsed variants), and the variant ID as the secondary key.
%The MapReduce partitioner uses custom Collect/sort functions to collect the tuples by their primary key (individual ID) and distributes them between nodes. The data is sorted by secondary key (variant ID)
%2. Reduce
%This data is input to the Reducers, which store the variants for each individual to a sparse vector with the Key as the individual ID and the value as the variant vector.
%
%Spark
%1. The lines from the VCF file(s) are parsed independently by worker nodes, where each line is zipped with the heading from the VCF file resulting in tuples of individual ID and the variant. 
%2. These arrays are each zipped again with the unique, sequential index (the Variant ID) resulting in tuples (essentially key/value pairs) of Variant IDs (?key?) and the array of tuples from step 1 (?value?).
%3. The flatmap function restructures the data into Individual ID (?key?) and tuples with Variant ID and the Variant (?value?).
%4. These tuples are grouped by the Individual ID, collecting all Variant ID and Variant tuples, and non-zero variants are saved as a sparse vector.
\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF file format to a data structure readable by Mahout and MLlib, respectively.}
      \end{figure}

>>>>

\subsection*{ADMIXTURE Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.

%% TODO - NEIL - (and better heading) %%
\subsection*{Personal Genome Project Data}
%% The PGP data..... 
PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
23andMe files were then converted to VCF using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
file using VCFtools \cite{Danecek2011Variant}.

@@

%Hadoop
%1. Map
%The mapper converts each line of the VCF file into a preliminary key-value pair, where the Key is the byte-position of the first character thereby functioning as a proxy for the genomic location (variant ID), and the Value is a string of the line. From this, the key-value pairs for the individual genotypes are created by extracting the variant from each sample as the new Value. The new key is a tuple consisting of the primary key, which is the individual ID (position of the parsed variants), and the variant ID as the secondary key.
%The MapReduce partitioner uses custom Collect/sort functions to collect the tuples by their primary key (individual ID) and distributes them between nodes. The data is sorted by secondary key (variant ID)
%2. Reduce
%This data is input to the Reducers, which store the variants for each individual to a sparse vector with the Key as the individual ID and the value as the variant vector.
%
%Spark
%1. The lines from the VCF file(s) are parsed independently by worker nodes, where each line is zipped with the heading from the VCF file resulting in tuples of individual ID and the variant. 
%2. These arrays are each zipped again with the unique, sequential index (the Variant ID) resulting in tuples (essentially key/value pairs) of Variant IDs (?key?) and the array of tuples from step 1 (?value?).
%3. The flatmap function restructures the data into Individual ID (?key?) and tuples with Variant ID and the Variant (?value?).
%4. These tuples are grouped by the Individual ID, collecting all Variant ID and Variant tuples, and non-zero variants are saved as a sparse vector.
\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF to a data structure readable by Mahout and MLlib, respectively.}
      \end{figure}
